Nail: A practical interface generator for data formats

### 2014-05-01 paper/osdi14/design.tex
DNS with line numbers

A Nail parser defines both the structure of some external format and a data type to represent that
format. Parsers are constructed by combinators over simpler parsers, an approach popularized by the Parsec
framework~\cite{LeijenMeijer:parsec}. We provide the most common combinators familiar from other parser
 combinator libraries, such as Parsec and Hammer~\cite{hammer-parser} and extend them so they also
 describe a data type. 
Figure~\ref{fig:syntax} shows the elements of Nail grammars, which we describe in more detail below.

\paragraph{Integers and constraints.}

Nail's fundamental parsers represent signed or unsigned
integers with arbitrary lengths up to 64 bits.
Note that is possible to define parsers for sub-byte lengths, e.g.,
to parse the 4-bit
data offset within the TCP header.  The internal representation rounds up to the nearest integer
type; for example, a 24-bit unsigned integer will be stored in an unsigned 32-bit integer.

@@

The \emph{many} combinator takes a parser and applies it repeatedly
until it fails, returning an array of the inner parser's results. The
\emph{sepBy} combinator
additionally takes a constant parser, which it applies in between parsing
two values, but not before parsing the first value or after parsing the
last.  This is useful for parsing an array of items delimited by a separator.
%For examle, \cc{many uint8} represents an array of 8-bit unsigned
%integers, and \cc{sepBy uint8=',' (many uint8 | '0'..'9')} recognizes
%comma-separated lists of decimal numbers.

@@

% \begin{figure}[tb]
% \smaller[0.5]
% \begin{minipage}{0.45\columnwidth}
% \begin{verbatim}
% dns_header = {
%   id uint16
%   qr uint1
%   opcode uint4
%   aa uint1 
%   tc uint1
%   rd uint1
%   ra uint1
%   uint3 = 0
%   rcode uint4
% }
% \end{verbatim} 
% \end{minipage}
% ~
% \begin{minipage}{0.45\columnwidth}
% \begin{verbatim}
% struct dns_header {
%   uint16_t id;
%   uint8_t qr;
%   uint8_t opcode;
%   uint8_t aa;
%   uint8_t tc;
%   uint8_t rd;
%   uint8_t ra;
%   uint8_t rcode;
% };
% \end{verbatim} 
% \end{minipage}

% \caption{Nail grammar (left) and data model (right) for a part of the
% grammar for DNS packets.  The \cc{uint3 = 0} grammar represents
% 3 bits of padding (filled with zeroes).}
% \label{fig:dns-struct}
% \end{figure}

Nail provides a structure combinator with semantic labels instead of the sequence combinator that
other parser combinator libraries use to capture structures in data formats. 
The structure combinator consists of a sequence of fields, typically consisting of a label and a
parser that describes the contents of that field. Other field types will be described below.
 The syntax of the structure combinator is inspired by the Go language~\cite{golang}, with field names preceding their definition.

@@

\paragraph{Constants.}
In some cases, not all bytes in a structure actually contain information, such as magic numbers or
reserved fields. Those fields can be represented in Nail grammars by constant fields in structures.
Constant fields do not correspond to a field in the internal model, but they are validated during
parsing and generated during output.  Constants can either have integer values, such as
\cc{uint3 = 0} or string values for text-based protocols, e.g. \cc{many uint8 = ``Foo''}.

@@

\paragraph{Wrap combinator.} 
When implementing real protocols with Nail, we often found
structures that consist of many constants and only one named field. This pattern is
common in binary protocols which use fixed headers to denote the type of data
structure to be parsed.  In order to keep the internal representation cleaner,
we introduced the wrap combinator, which takes a sequence of parsers containing
exactly one non-constant parser. The parser and generator act as though the wrap
combinator was a sequence of parsers, but the data model does not wrap the
single value in another structure, making the application-visible representation
(and thus application code) more concise.
For example, \cc{<uint8='"'; many int8 | 'a'..'z'; uint8='"'>} parses a quoted
lowercase word into an array.

@@

\paragraph{Optional.}
Nail includes an \cc{optional} combinator, which attempts to recognize a value, but succeeds
without consuming input when it cannot recognize that value. Syntactically, \cc{optional} is
equivalent to a choice between the parser and an empty structure, but in the internal model it is
more concisely represented as a reference that is null when the parser fails.
For example, the grammar for  Ethernet headers uses \cc{optional vlan\_header} to parse the VLAN
header that only appears in Ethernet packets transmitted to a non-default VLAN.
\paragraph{Rules.}
A Nail grammar consists of rules that assign a parser to a name. Rules are written as
assignments, such as \cc{ints = many uint8}, which defines a rule called \cc{ints} with the
body \cc{many uint8}.  As we will describe in \S\ref{s:dependent}
and \S\ref{s:transforms}, rules can optionally consume parameters. 
Rules can be invoked in a Nail grammar anywhere a parser can appear. Rule invocations act as though
the body of the rule had been substituted in the code. If parameters appear, they are passed by reference.

>>>>

A Nail parser defines both the structure of some external format and a data type to represent that
format. Parsers are constructed by combinators over simpler parsers, an approach popularized by the Parsec
framework~\cite{LeijenMeijer:parsec}. We provide the most common combinators familiar from other parser
 combinator libraries, such as Parsec and Hammer~\cite{hammer-parser} and extend them so they also
 describe a data type.

We present both a systematic overview of Nail's syntax with short examples in
Figure~\ref{fig:syntax} shows the elements of Nail grammars and explain our design in more detail
below, using a grammar for the well-known DNS protocol as a guiding example.

\input{fig-dnsfull}
\paragraph{Rules.}
A Nail grammar consists of rules that assign a parser to a name. Rules are written as assignments,
such as \cc{ints = /*parser definition*/}, which defines a rule called \cc{ints}. As we will
describe later in \S\ref{s:dependent} and \S\ref{s:transforms}, rules can optionally consume parameters.
Rules can be invoked in a Nail grammar anywhere a parser can appear. Rule invocations act as though
the body of the rule had been substituted in the code. If parameters appear, they are passed by
reference.
\paragraph{Integers and constraints.}

Nail's fundamental parsers represent signed or unsigned
integers with arbitrary lengths up to 64 bits.
Note that is possible to define parsers for sub-byte lengths, for example, the flag bits in the DNS
message header, in lines 4 through 8.

@@

The \emph{many} combinator takes a parser and applies it repeatedly
until it fails, returning an array of the inner parser's results. In line 39 of the DNS grammar, a
sequence of labels is parsed by parsing as many labels as possible, that is, until an invalid length
field is encountered.
The \emph{sepBy} combinator
additionally takes a constant parser, which it applies in between parsing
two values, but not before parsing the first value or after parsing the
last.  This is useful for parsing an array of items delimited by a separator.
%For examle, \cc{many uint8} represents an array of 8-bit unsigned
%integers, and \cc{sepBy uint8=',' (many uint8 | '0'..'9')} recognizes
%comma-separated lists of decimal numbers.

@@

Nail provides a structure combinator with semantic labels instead of the sequence combinator that
other parser combinator libraries use to capture structures in data formats. 
The structure combinator consists of a sequence of fields, typically consisting of a label and a
parser that describes the contents of that field, surrounded by curly braces. Other field types will be described below.
 The syntax of the structure combinator is inspired by the Go language~\cite{golang}, with field
 names preceding their definition.

@@

\paragraph{Constants.}
In some cases, not all bytes in a structure actually contain information, such as magic numbers or
reserved fields. Those fields can be represented in Nail grammars by constant fields in structures.
Constant fields do not correspond to a field in the internal model, but they are validated during
parsing and generated during output.  Constants can either have integer values, such as
in line 9 of the DNS grammar, or string values for text-based protocols, e.g. \cc{many uint8 = ``Foo''}.

@@

\paragraph{Wrap combinator.} 
When implementing real protocols with Nail, we often found
structures that consist of many constants and only one named field. This pattern is
common in binary protocols which use fixed headers to denote the type of data
structure to be parsed.  In order to keep the internal representation cleaner,
we introduced the wrap combinator, which takes a sequence of parsers containing
exactly one non-constant parser. The external format is defined as though the wrap combinator were a
structure, but the data model does not introduce a structure with just one element, making the application-visible representation
(and thus application code) more concise.
Line 39 of the DNS grammar uses the wrap combinator to hide the terminating NUL-byte of a sequence
of labels.

@@

\paragraph{Optional.}
Nail includes an \cc{optional} combinator, which attempts to recognize a value, but succeeds
without consuming input when it cannot recognize that value. Syntactically, \cc{optional} is
equivalent to a choice between the parser and an empty structure, but in the internal model it is
more concisely represented as a reference that is null when the parser fails.
For example, the grammar for  Ethernet headers uses \cc{optional vlan\_header} to parse the VLAN
header that only appears in Ethernet packets transmitted to a non-default VLAN.

### 2014-05-01 paper/osdi14/eval.tex
Eval
DNS with line numbers

Figure~\ref{fig:dns-xform} presents the Nail grammar for DNS packets.
The grammar corresponds almost directly to the diagrams in RFC 1035,
which defines DNS~\cite[\S4]{RFC:1035}.  Each DNS packet consists of a
header, a set of question records, and a set of answer records. Domain
names in both queries and answers are encoded as a sequence of labels,
terminated by a zero byte.  Labels are Pascal-style strings, consisting
of a length field followed by that many bytes comprising the label.

@@

\input{fig-dnsfull}

@@

An especially tricky data format is the ZIP compressed archive
format~\cite{pkzip}.  ZIP files are normally parsed end-to-beginning. At
the end of each ZIP file is an \emph{end-of-directory header}. This header
contains a variable-length comment, so it has to be located by scanning
backwards from the end of the file until a magic number and a valid
length field is found. Many ZIP implementations disagree on how to find
this header in confusing situations, such as when the comment contains
the magic number~\cite{wolf:berlinsides-zip}.  This end-of-directory
header contains the offset and size of the \emph{ZIP directory}, which
is an array of \emph{directory entry headers}, one for every file
in the archive.  Each entry stores file metadata, such as file name,
compressed and uncompressed size, and a checksum, in addition to the
offset of a \emph{local file header}. The local file header duplicates
most information from the directory entry header and the compressed file
contents follow it immediately.

@@

The \cc{zip\_file} grammar first splits the entire file stream into two
streams based on the \cc{zip\_end\_of\_directory} transform (whose two
C function signatures are shown in Figure~\ref{fig:zip-eod-xform}).
\cc{zip\_end\_of\_directory_parse} finds the end-of-directory header
as described above, by scanning the file backwards, and splits the
file into two streams, one containing the end-of-directory header and
one containing the file contents.  The header stream is then parsed,
according to the \cc{apply} rule, using the \cc{end\_of\_directory}
parser.  The file stream, to which all the offsets in the metadata refer,
is passed as an argument to that parser.  The directory header is then
parsed, and the offset and size transforms provided by Nail are used
to isolate the actual directory from the file stream.  Each directory
entry in turn points to a local file header.

The \cc{file} grammar for a ZIP file entry is particularly interesting.
It receives as arguments the file size, checksum, and compression method
of the file from the directory header.  However, this same information
is duplicated in the file entry, so the grammar uses the Nail-supplied
\cc{u16\_depend} transform to check whether the two values are equal. Unlike
most other transforms, \cc{u16\_depend} does not consume or produce strings; it
only checks that two dependent fields are equal when parsing, and assigns
the value of the second field to the first when generating.  This ensures
that the programmer does not have to worry about inconsistencies when
handling the internal representation of a ZIP file.

>>>>

In Section~\ref{s:design}, we introduced Nail's syntax with a grammar
for DNS packets, shown in Figure~\ref{fig:dns-xform}.
The grammar corresponds almost directly to the diagrams in RFC 1035,
which defines DNS~\cite[\S4]{RFC:1035}.  Each DNS packet consists of a
header, a set of question records, and a set of answer records. Domain
names in both queries and answers are encoded as a sequence of labels,
terminated by a zero byte.  Labels are Pascal-style strings, consisting
of a length field followed by that many bytes comprising the label.

@@

@@

An especially tricky data format is the ZIP compressed archive
format~\cite{pkzip}.  ZIP files are normally parsed end-to-beginning. At
the end of each ZIP file is an \emph{end-of-directory header}. This header
contains a variable-length comment, so it has to be located by scanning
backwards from the end of the file until a magic number and a valid
length field is found. Many ZIP implementations disagree on how to find
this header in confusing situations, such as when the comment contains
the magic number~\cite{wolf:berlinsides-zip}.

 This end-of-directory
header contains the offset and size of the \emph{ZIP directory}, which
is an array of \emph{directory entry headers}, one for every file
in the archive.  Each entry stores file metadata, such as file name,
compressed and uncompressed size, and a checksum, in addition to the
offset of a \emph{local file header}. The local file header duplicates
most information from the directory entry header. The compressed file
contents follow the header immediately.

@@

The \cc{zip\_file} grammarfirst splits the entire file stream into two
streams based on the \cc{zip\_end\_of\_directory} transform in line 2
(whose two C function signatures are shown in
Figure~\ref{fig:zip-eod-xform}). \cc{zip\_end\_of\_directory_parse}
finds the end-of-directory header as described above, by scanning the
file backwards, and splits the file into two streams, one containing
the end-of-directory header and one containing the file contents. The
\cc{end\_of\_directory} rule is then applied to the header stream in
line 4. All offsets in the ZIP file refer to the beginning of the
file, so the stream \cc{\$file} which contains the file contents
without the header is passed as an argument to all parsers from
hereon.

The directory header contains the offset and size of the ZIP
directory (lines 9 and 10). The \cc{offset} and \cc{size}
transformations extract a stream containing just the directory from
the file contents. This stream is then parsed as an array of directory
entries in line 17.
Each directory entry in turn points to a local file header, which is
similarly extracted and parsed with the \cc{file} rule.

The \cc{file} rule starting at line 41, describing a ZIP file entry takes dependent field
parameters containing file metadata information from the directory
header. However, this same information is duplicated in the file
entry, so the grammar uses the Nail-supplied \cc{u16\_depend}
transform to check whether the two values are equal. Unlike most other
transforms, \cc{u16\_depend} does not consume or produce strings; it
only checks that two dependent fields are equal when parsing, and
assigns the value of the second field to the first when generating.
This ensures that the programmer does not have to worry about
inconsistencies when handling the internal representation of a ZIP
file.
